---
title: "02_sunrisetimes"
author: "Michaela Gustafson"
date: "12/30/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Sunrise Times for Nome, AK

### Description:
Import and clean sunrise times for Nome, AK to be merged with environmental variables data frame
Sunrise times are from dateandtime.com

## LIBRARY
Packages used for this code:

```{r library}
library(here) # for file location
library(hms) # for making column into a time format
library(tidyr) # general cleaning/manipulation
library(dplyr) # general cleaning/manipulation
```

## IMPORT SUNRISE DATA

#### 2019 Sunrise Times

```{r 19sun}
sun19 <- read.csv(here("data/sunrise_2019.csv"))
```

#### 2021 Sunrise Times

```{r}
sun21 <- read.csv(here("data/sunrise_2021.csv"))
```


## CLEAN SUNRISE DATA

First need add extra 0 to beginning of those with one less character so column 
can be formatted in hour:min:sec

### 2019 Sunrise Times

```{r}
# remove colons(:) first so that leading zero can be added
sun19$sunrise <- base::gsub(":", "", sun19$sunrise) 

# add leading 0
sun19$sunrise <- base::sprintf("%06d", as.numeric(sun19$sunrise)) 

# separate sunrise times into columns for h:m:s
sun19 <- tidyr::extract(sun19, 
                        sunrise, 
                        into = c("hr", "min", "sec"), 
                        "(.{2})(.{2})(.{2})", 
                        remove=FALSE)

# paste h:m:s column together into another column and separate with ':'
sun19$sunrise <- base::paste(sun19$hr, sun19$min, sun19$sec, sep=":")

# format column so that R recognizes it as time
sun19$sunrise <- hms::as_hms(sun19$sunrise)

#check structure
str(sun19)

# keep only date and sunrise times in the df:
colnames(sun19)
sun19 <- sun19[c(1, 2)]

# write final sunrise times as a .csv
write.csv(sun19, here("data/nome_sun19_final.csv"))

```

### 2021 Sunrise Times 

```{r}
# remove colons in time to be able to add leading 0
sun21$sunrise <- base::gsub(":", "", sun21$sunrise) 
head(sun21)

# addextra 0 to 5chr strings of time so column can be formatted in h:m:s

sun21$sunrise <- base::sprintf("%06d", as.numeric(sun21$sunrise)) 

# separate sunrise times into columns for h:m:s
sun21 <- tidyr::extract(sun21, 
                        sunrise, 
                        into = c("hr", "min", "sec"), 
                        "(.{2})(.{2})(.{2})", 
                        remove=FALSE)

# paste h:m:s column together into another column and separate with ':'
sun21$sunrise <- base::paste(sun21$hr, sun21$min, sun21$sec, sep=":")

# format column so that R recognizes it as time
sun21$sunrise <- hms::as.hms(sun21$sunrise)

#check structure
str(sun21)


# keep only date and sunrise times in the df:
colnames(sun21)
sun21 <- sun21[c(1, 2)]

# write final sunrise times as a .csv
write.csv(sun21, here("data/nome_sun21_final.csv"))

```


## END SCRIPT 



