---
title: "06_addelevation"
author: "Michaela Gustafson"
date: "1/3/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Adding Elevation to Environmental Data

In this code, I will be running a focal function over a DEM (digital elevation model) of the study are on the Seward Peninsula to calculate the mean elevation within a 390m buffer. From the focal(ized) raster, I will extract mean elevation at each survey point. 


## LIBRARY
These are the packages used in this code:
```{r library}
library(terra)
library(raster)
library(sf)
library(sp)
library(here)
library(ggplot2)
library(tidyverse)
library(raster)
library(taRifx)
library(tidyr)
library(dplyr)
library(stringr)

```


## IMPORT DEM

```{r dem}
ak_dem <- rast(here("data/DEM_Michaela/DEM_Michaela/dsm_mosaic_AK.tif"))
plot(ak_dem)

crs(ak_dem)

```

## IMPORT SURVEY POINTS

```{r importpoints}
# 2019
points19 <- read.csv(here("data/points_2019_all_FINAL.csv"))
## make shapefile and assign crs
points19.sf <- st_as_sf(points19, 
                        coords = c("lon", "lat"),
                        crs = 4326)

plot(st_geometry(points19.sf))

st_crs(points19.sf)

st_is_valid(points19.sf)

# 2021

points21 <- read.csv(here("data/utms_21_final.csv"))
points21 <- points21[,c(2:4)]

# need to assign crs
# points were taken in UTM UPS, NAD 83, GRS 80
# epsg code: https://epsg.io/26903

points21.sf <- st_as_sf(points21, 
                         coords = c("UTM1", "UTM2"),
                         crs = 26903)

st_crs(points21.sf)
plot(points21.sf)

st_is_valid(points21.sf)



```

### Make sure in same CRS as DEM
Make sure that CRS of point data matches with elevation raster

```{r reproject}

# reproject points to same crs as hab raster

#st_transform

# 2019 points
points19.proj <- st_transform(points19.sf, crs(ak_dem))
st_crs(points19.proj)
plot(points19.proj)

# reproject 2021 points to match hab raster crs
points21.proj <- st_transform(points21.sf, crs(ak_dem))
st_crs(points21.proj)
plot(points21.proj)


# check
plot(ak_dem)
plot(points19.proj, add = TRUE)
plot(points21.proj, add = TRUE)
```

## FOCAL FOR MEAN ELEVATION
```{r focal}
?terra::focal
ak.dem.focal <- terra::focal(ak_dem, w = matrix(1, nc=13, nr=13), fun = mean, na.rm = TRUE)
plot(ak_dem)
plot(ak.dem.focal)

# create raster of focalized dem

writeRaster(ak.dem.focal, filename=file.path(here(), "ak_dem_focal.tif"), overwrite = TRUE)

```
## EXTRACT ELEVATION AT EACH POINT
Now we will extract the elevation at each survey point to get the mean elevation within a 390m buffer (this comes from extracting from the focal raster)

```{r extract}
?terra::extract

# first need to turn points into SpatVectors

points19.vect <- terra::vect(points19.proj)
points21.vect <- terra::vect(points21.proj)

elev.19 <- terra::extract(ak.dem.focal, points19.vect)
elev.21 <- terra::extract(ak.dem.focal, points21.vect)
```


### REJOIN WITH ORIGINAL IDs

```{r join}
elev19.join <- as.data.frame(cbind(points19.proj, elev.19))
elev21.join <- as.data.frame(cbind(points21.proj, elev.21))
```


### CLEAN UP
Clean up the final dataframe. Change eleveation column name, remove unecessary id column.

```{r clean}
# 2019

colnames(elev19.join)
elev19final <- elev19.join[,c(2,4)]
# check
head(elev19final)

# rename elevation column
elev19final <- rename(elev19final, elev = dsm_mosaic_AK)

# check
head(elev19final)

# 2021
colnames(elev21.join)
# select only the columns I want to keep
elev21final <- elev21.join[,c(1, 3)]
# check
head(elev21final)

# change elevation column name
elev21final <- rename(elev21final, elev = dsm_mosaic_AK)

# check
head(elev21final)
```

#### Export elevation files

```{r export}
write.csv(elev19final, here("data/mean_elev_19.csv"))
write.csv(elev21final, here("data/mean_elev_21.csv"))
```















